Staged Abstract Interpreters

### 2019-09-08 paper/50_stageabs.tex
typo

The trait @RepLattice@ is shown by specializing @Lattice@ with type @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
are eventually delegated to the operations on a staged set data type, such as
@subsetOf@, @union@ and @intersect@. We do not need to change the
implementation code, except the types -- such as from @Set[T]@ to @Rep[Set[T]]@.
We show the changes in light gray code. The LMS library provides implicit
conversions that lift the current-stage values to their next-stage values. In
the code generation part, these operations emit their corresponding next-stage
code. Again, other lattice structures such as products and maps are implemented
in a similar way.

@@

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that the recursive calls of the evaluator @ev@ on the lambda body are shifted from
@ap_clo@ to @close@. In other words, we eagerly specialize the interpreter and
generate code every time when we see a $\lambda$ term, instead of lazily when the
applications happen.  With no surprise, it is similar here. Additionally,
in the staged abstract interpreter, we have to handle nondeterminism
incurred by the over-approximation of runtime behavior.
In the rest of this part, we keep our roadmap and focus on discussing closure
representations and function applications for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the @Ans@ monadic value to
a grounded value of type @Result@, by providing the desired arguments, i.e., the
new environment, new store, and caches. The collapsing of monad happens at
current-stage, so the calls of @ev@ is unfolded within staging. Finally, we
generate a singleton set containing the compiled next-stage closure
@emit_compiled_clo(f)@, which is represented by an IR node in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure @clo@ in the set, by using @emit_ap_clo@. The
method @emit_ap_clo@ generates the current-stage representation of the
future-stage application result.
Finally, we put the future @out@ cache, store, and values back into
the current-stage monads.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(funs: Rep[Value], arg: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](funs)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

\paragraph{Staged Caching and Fixpoint Iteration} Our fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@. Consequently, the branching condition cannot be determined
statically -- we need to generate code for the whole @if@ expression. Figure
\ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the queried result form
the @out@ cache and the @out@ cache itself. The else branch constructs a monadic
value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@ cache.
After which, we use a similar technique that collapses the monadic value @m@ to
grounded values, by providing its desired environment and etc. Finally, we have a
current-stage representation of the future-stage values @res@, and put the
content of @res@ back into the monad stack.

>>>>

The trait @RepLattice@ is shown by instantiating @Lattice@ with type @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
are eventually delegated to the operations on a staged set data type, such as
@subsetOf@, @union@ and @intersect@. We do not need to change the
implementations, except the types -- such as changing from @Set[T]@ to
@Rep[Set[T]]@.  We show the changes in light gray code. The LMS library
provides implicit conversions that lift the current-stage values to their
next-stage representations. In the code generation part, these operations emit
their corresponding next-stage code. Again, other lattice structures such as
products and maps are implemented in a similar way.

@@

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that the recursive calls of the evaluator @ev@ on the lambda body are shifted from
@ap_clo@ to @close@. In other words, we eagerly specialize the interpreter and
generate code every time when we see a $\lambda$ term, instead of lazily
calling @ev@ when the applications happen.  With no surprise, it is similar
here. Additionally, in the staged abstract interpreter, we have to handle
nondeterminism incurred by the over-approximation of runtime behavior.  In the
rest of this part, we keep our roadmap and focus on discussing closure
representations and function applications for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the monadic value of
type @Ans@ to grounded values of type @Result@, by providing the desired
arguments, i.e., the new environment, new store, and caches. The collapsing of
monads happens at the current-stage, so the calls of @ev@ is unfolded within
staging. Finally, we generate a singleton set containing the compiled
next-stage closure @emit_compiled_clo(f)@, which is represented by an IR node
in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure @clo@ in the set, by using @emit_ap_clo@. The
method @emit_ap_clo@ generates the current-stage representation of the
future-stage application result.
Finally, we reify the future @out@ cache, store, and values back into
the current-stage monads.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(funs: Rep[Value], arg: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](funs)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

\paragraph{Staged Caching and Fixpoint Iteration} Our fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Because of this, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@. The branching condition cannot be determined
statically, consequently we need to generate code for the whole @if@ expression. Figure
\ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the queried result
form the @out@ cache and the @out@ cache itself. The else branch constructs a
monadic value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@
cache.  After which, we use a similar technique that collapses the monadic
value @m@ to grounded values, by providing its desired environment and etc.
Finally, we have a current-stage representation of the future-stage values
@res@, and put the content of @res@ back into the monad stack.
